Model-based pretreatment for regression with correlated variables.

### 2014-04-16 www/article13.tex
up
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/correg@543 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

We want to find the most probable structure $S$ knowing the dataset, so we search for the structure that maximizes $P(S|\boldsymbol{X})$ and we have:	
	\begin{eqnarray}
	 \label{approxBIC} P(S|\boldsymbol{X})&\propto & P(\boldsymbol{X}|S)P(S)
	=P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)P(\boldsymbol{X}_f|S)P(S)
	\end{eqnarray}

	The discrete structure $S$ is identifiable because we can't permute some regressions in (\ref{SR}) and obtain the same joint distribution $P(\boldsymbol{X},\boldsymbol{Y})$, the residuals ($\boldsymbol{\varepsilon}_j$) would not stay Gaussian (see the appendices).

	\subsection{Modeling the uncorrelated covariates: a full generative approach on $P(\boldsymbol{X})$}
	To be able to compare structures with probabilistic likelihood-based criterions, we need a full generative model on $\boldsymbol{X}$. Sub-regressions give $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S) $ but $P(\boldsymbol{X}_f|S)$ is still undefined. We suppose that variables in $\boldsymbol{X}_f$ follow Gaussian mixtures of $k_j \in \mathbf{N}^*$ components: 
	\begin{equation}
			\forall j \notin I_r : \boldsymbol{X}^j_{|S} \sim f(\boldsymbol{\theta}_j)=\mathcal{GM}(\boldsymbol{\pi}_j;\boldsymbol{\mu}_j;\boldsymbol{\sigma}^2_j) \textrm{ with } \boldsymbol{\pi}_j,\boldsymbol{\mu}_j,\boldsymbol{\sigma}^2_j \textrm{ vectors of size } k_j. \label{mixtureX1}
		\end{equation}
		The great flexibility \cite{mclachlan2004finite} of such models makes our model more robust but one can use other laws if needed (exponentially distributed for example). Gaussian case is just a special case ($k_j=1$) of Gaussian mixture so it is included in our hypothesis.

@@

	We note $\boldsymbol{\Theta}$ the set of the parameters of the generative model
	\begin{eqnarray}
		-2\log P(\boldsymbol{X}|S)&\approx & BIC=-2\mathcal{L}(\boldsymbol{X},S,\boldsymbol{\Theta})+|\boldsymbol{\Theta}|\log(n)  
	\end{eqnarray}
	But $BIC$ tends to give too complex structures because we test a great range of models. 
	Thus we choose to penalise the complexity a bit more with a hierarchical uniform {\it a priori} distribution $P(S)=P(I_f | \boldsymbol{p}_f,I_r,p_r)P(\boldsymbol{p}_f|I_r,p_r)P(I_r|p_r)P(p_r)$  instead of a simple uniform law on $S$ that is generally used and provides no penalty.
%	 Thus we have :
%		\begin{eqnarray}
%		BIC^*(X|S)&=&BIC(X|S) +\ln(P(S)) \label{Bicstar}
%	\end{eqnarray}		
	It increases penalty on complexity for $p_r<\frac{p}{2}$ and $p_f^j<\frac{p}{2}$ . Hence %when using $BIC*$ 
	this constraint on $\hat{p}_r$ and $\hat{p}_f^j$ is given in the research algorithm when the Hierarchical Uniform (HU) hypothesis is made instead of Uniform one (U) in numerical experiments (section \ref{sectionsimul} and \ref{sectionrealcase}).
		We do not change $BIC$ but only $P(S)$ so the properties are the same as classical $BIC$ but we obtain better results when (HU) this constraint is verified.

@@

	The walk follows a time-homogeneous Markov Chain whose transition matrix $\mathcal{P}$ has $|\mathcal{S}|$ rows and columns (combinatory so we just compute the probabilities when we need them).
	At each step the markov chain moves with probabiliy:
	\begin{eqnarray}
			\forall (S,\tilde{S}) \in \mathcal{S}^2 : \mathcal{P}(S,\tilde{S})&=&\sum_{j=1}^p \mathbf{1}_{ \{\tilde{S}\in \mathcal{V}_{S,j}\} }\frac{\exp(-\frac{1}{2}P(\tilde{S}|\boldsymbol{X}))}{\sum_{S_l\in \mathcal{V}_{S,j}}\exp(-\frac{1}{2}P(S_l|\boldsymbol{X}))} 
	\end{eqnarray}
	And $\mathcal{S}$ is a finite state space.%la relaxation rend P non symétrique mais ne remets  pas en cause l'homogénéité

@@

 \subsubsection{Initialisation(s)}
 If we have some knowledge about some sub-regressions (physical models for example) we can add them in the found and/or initial structure. So the model is really expert-friendly.
The initial structure can be based on a first warming algorithm taking the correlations into account. coefficients are randomly placed into $I_1$, weighted by the absolute value of the correlations. We do so in the followings. Then this structure could be for example reduced by the hadamard product with the binary matrix obtained by Graphical Lasso\cite{friedman2008sparse} that makes selection in the precision matrix but it is time consuming.

	One would rather test multiple short chains than lose time in initialisation or long chains \cite{gilks1996markov}. It also helps to face local extrema. In the followings, the chain was launched with twenty initialisations.

>>>>

We want to find the most probable structure $S$ knowing the dataset, so we search for the structure that maximizes $P(S|\boldsymbol{X})$ and we have:	
	\begin{eqnarray}
	 \label{approxBIC} P(S|\boldsymbol{X})&\propto & P(\boldsymbol{X}|S)P(S)
	=P(\boldsymbol{X}_r|\boldsymbol{X}_f,S)P(\boldsymbol{X}_f|S)P(S)
	\end{eqnarray}
So we will try to maximize $\psi(\boldsymbol{X},S)=P(\boldsymbol{X}|S)P(S)$.
	The discrete structure $S$ is identifiable because we can't permute some regressions in (\ref{SR}) and obtain the same joint distribution $P(\boldsymbol{X},\boldsymbol{Y})$, the residuals ($\boldsymbol{\varepsilon}_j$) would not stay Gaussian (see the appendices).

	\subsection{Modeling the uncorrelated covariates: a full generative approach on $P(\boldsymbol{X})$}
	To be able to compare structures with $P(S|\boldsymbol{X})$, we need a full generative model on $\boldsymbol{X}$. Sub-regressions give $P(\boldsymbol{X}_r|\boldsymbol{X}_f,S) $ but $P(\boldsymbol{X}_f|S)$ is still undefined. We suppose that variables in $\boldsymbol{X}_f$ follow Gaussian mixtures of $k_j \in \mathbf{N}^*$ components: 
	\begin{equation}
			\forall j \notin I_r : \boldsymbol{X}^j_{|S} \sim f(\boldsymbol{\theta}_j)=\mathcal{GM}(\boldsymbol{\pi}_j;\boldsymbol{\mu}_j;\boldsymbol{\sigma}^2_j) \textrm{ with } \boldsymbol{\pi}_j,\boldsymbol{\mu}_j,\boldsymbol{\sigma}^2_j \textrm{ vectors of size } k_j. \label{mixtureX1}
		\end{equation}
		The great flexibility \cite{mclachlan2004finite} of such models makes our model more robust but one can use other laws if needed (exponentially distributed for example). Gaussian case is just a special case ($k_j=1$) of Gaussian mixture so it is included in our hypothesis.

@@

	We note $\boldsymbol{\Theta}$ the set of the parameters of the generative model
	\begin{eqnarray}
		-2\log P(\boldsymbol{X}|S)&\approx & BIC=-2\mathcal{L}(\boldsymbol{X},S,\boldsymbol{\Theta})+|\boldsymbol{\Theta}|\log(n)  
	\end{eqnarray}
	But $BIC$ tends to give too complex structures because we test a great range of models. 
	Thus we choose to penalise the complexity a bit more with a hierarchical uniform {\it a priori} distribution $P(S)=P(I_f | \boldsymbol{p}_f,I_r,p_r)P(\boldsymbol{p}_f|I_r,p_r)P(I_r|p_r)P(p_r)$  instead of a simple uniform law on $S$ that is generally used and provides no penalty.
%	 Thus we have :
%		\begin{eqnarray}
%		BIC^*(X|S)&=&BIC(X|S) +\ln(P(S)) \label{Bicstar}
%	\end{eqnarray}		
	It increases penalty on complexity for $p_r<\frac{p}{2}$ and $p_f^j<\frac{p}{2}$ . Hence %when using $BIC*$ 
	this constraint on $\hat{p}_r$ and $\hat{p}_f^j$ is given in the research algorithm when the Hierarchical Uniform (HU) hypothesis is made instead of Uniform one (U) in numerical experiments (section \ref{sectionsimul} and \ref{sectionrealcase}).
		We do not change $BIC$ but only $P(S)$ so the properties are the same as classical $BIC$ but we obtain better results when (HU) this constraint is verified. Uniform law on $P(S)$ gives $\psi(\boldsymbol{X},S)=P(\boldsymbol{X}|S)$ so it is equivalent to a maximization of the $BIC$. With the Hierarchical Uniform hypothesis we maximize $\psi(\boldsymbol{X},S)\approx BIC + P(S)$.

@@

	The walk follows a time-homogeneous Markov Chain whose transition matrix $\mathcal{P}$ has $|\mathcal{S}|$ rows and columns (combinatory so we just compute the probabilities when we need them).
	At each step the markov chain moves with probabiliy:
	\begin{eqnarray}
			\forall (S,\tilde{S}) \in \mathcal{S}^2 : \mathcal{P}(S,\tilde{S})&=&\sum_{j=1}^p \mathbf{1}_{ \{\tilde{S}\in \mathcal{V}_{S,j}\} }\frac{\exp(-\frac{1}{2}\psi(\boldsymbol{X},\tilde{S}))}{\sum_{S_l\in \mathcal{V}_{S,j}}\exp(-\frac{1}{2}\psi(\boldsymbol{X},S_l))} 
	\end{eqnarray}
	And $\mathcal{S}$ is a finite state space.%la relaxation rend P non symétrique mais ne remets  pas en cause l'homogénéité

@@

 \subsubsection{Initialisation(s)}
 If we have some knowledge about some sub-regressions (physical models for example) we can add them in the found and/or initial structure. So the model is really expert-friendly.
The initial structure can be based on a first warming algorithm taking the correlations into account. Coefficients are randomly placed into $I_f$, weighted by the absolute value of the correlations. We do so in the followings. Then this structure could be for example reduced by the Hadamard product with the binary matrix obtained by Graphical Lasso\cite{friedman2008sparse} that makes selection in the precision matrix but it is time consuming.

	One would rather test multiple short chains than lose time in initialisation or long chains \cite{gilks1996markov}. It also helps to face local extrema. In the followings, the chain wzs launched with twenty initialisations each time.
